Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-13 paper/main.tex
minor
add knuth morris pratt
new total number of lines analyzed

\begin{abstract}
  This paper presents a new approach for automatically deriving
  worst-case resource bounds for C programs.
  %
  The described technique combines ideas from amortized analysis and
  abstract interpretation in a unified framework to address four
  challenges for state-of-the-art techniques: compositionality, user
  interaction, generation of proof certificates, and scalability.
  %
  \emph{Compositionality} is achieved by incorporating the potential
  method of amortized analysis.  It enables the derivation of global
  whole-program bounds with local derivation rules by naturally
  tracking size changes of variables in sequenced loops and function
  calls.  The resource consumption of functions is described abstractly
  and a function call can be analyzed without access to the function
  body.
  %\
  \emph{User interaction} is supported with a new mechanism that
  clearly separates qualitative and quantitative verification. A user
  can guide the analysis to derive complex non-linear bounds by using
  auxiliary variables and assertions.  The assertions are
  separately proved using established qualitative techniques such as
  abstract interpretation or Hoare logic.
  %
  \emph{Proof certificates} are automatically generated from the local
  derivation rules. A soundness proof of the derivation system with
  respect to a formal cost semantics guarantees the validity of the
  certificates.
  %
  \emph{Scalability} is attained by an efficient reduction of
  bound inference to a linear optimization problem that can be solved
  by off-the-shelf LP solvers.
  %
  The analysis framework is implemented in the publicly-available
  tool $\toolname$. An experimental evaluation demonstrates the
  advantages of the new technique with a comparison of $\toolname$ with
  existing tools on challenging micro benchmarks and the analysis of
  more than 2000 lines of C code from the cBench benchmark suite.

@@

We have implemented the analysis system in the tool
\toolname and experimentally evaluated its effectiveness by
analyzing system code and examples from the literature.
\toolname has automatically derived global resource bounds for
more than 2000 lines of C code from the cBench benchmark suite.
%
\iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare $\toolname$
with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.

@@

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that can
  become available during execution.
\itemskipIn
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\itemskipIn
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\itemskipIn
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\itemskipIn
\item We implemented our resource bound analysis in the publicly-available tool $\toolname$.
\itemskipIn
\item We present experiments with $\toolname$ on more
  than 2000 lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip
%
\iffull{}
{\footnotesize \sl
[To the reviewers: The auxiliary material contains an anonymized
version of the extended article, and the source code of $\toolname$
and all examples.] }

@@

\pref{tab:cbench} contains a compilation of the results of our
experiments with the cBench benchmark suite. It shows a representative
list of automatically derived function bounds. In total we analyzed
more than 2000 lines of code.  In the LoC column
we not only count the lines of the analyzed function but
also the ones of all the function it calls. \iffull{The line count is
computed on pre-processed C with all unnecessary declarations removed.}{}
We analyzed the functions using a metric that assigns a cost 1 to all
the back-edges in the control flow (loops, and function calls).
The bounds for the functions \code{ycc\_rgb\_conv} and \code{uv\_decode}
have been inferred with user interaction as described in \pref{sec:anno}.
The most challenging functions for $\toolname$ have unrolled loops
where many variables are assigned.  This stresses our analysis because
the number of LP variables has a quadratic growth in program variables.
Even on these stressful examples, the analysis could finish in less
than 2 seconds.  \iffull{These times can be reduced by adding
manual annotations to indicate $\toolname$ that certain variables do
not have to be tracked.  We could also envision to implement heuristics.}{}
%
For example, the \code{sha\_update} function
is composed of one loop calling two helper functions that in turn
have 6 and 1 inner loops.  In the analysis of the SHA algorithm,
the compositionality of our analysis is essential to get a tight bound
since loops on the same index are sequenced 4 and 2 times without
resetting it.  All other tools derive much larger constant factors.

@@

\iffull{
\begin{table}[t]
\centering
\small
\begin{tabular}{r|c|c|c}
Function & LoC & Bound & Time (s) \\
\hline
\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
\code{MD5Final} & 195 & 141 & 0.22 \\
\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
\hline
\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
\end{tabular}
\vspace{.1cm}
\caption{Derived bounds for functions from cBench.}
\label{tab:cbench}
\end{table}
}
{
\begin{table}[t]
\small
\centering
\begin{tabular}{r|c|c|c}
Function & LoC & Bound & Time (s) \\
\hline
\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
\hline
\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
\end{tabular}
\vspace{.1cm}
\caption{Derived bounds for functions from cBench.}
\label{tab:cbench}
\end{table}
}

>>>>

\begin{abstract}
  This paper presents a new approach for automatically deriving
  worst-case resource bounds for C programs.
  %
  The described technique combines ideas from amortized analysis and
  abstract interpretation in a unified framework to address four
  challenges for state-of-the-art techniques: compositionality, user
  interaction, generation of proof certificates, and scalability.
  %
  \emph{Compositionality} is achieved by incorporating the potential
  method of amortized analysis.  It enables the derivation of global
  whole-program bounds with local derivation rules by naturally
  tracking size changes of variables in sequenced loops and function
  calls.  The resource consumption of functions is described abstractly
  and a function call can be analyzed without access to the function
  body.
  %\
  \emph{User interaction} is supported with a new mechanism that
  clearly separates qualitative and quantitative verification. A user
  can guide the analysis to derive complex non-linear bounds by using
  auxiliary variables and assertions.  The assertions are
  separately proved using established qualitative techniques such as
  abstract interpretation or Hoare logic.
  %
  \emph{Proof certificates} are automatically generated from the local
  derivation rules. A soundness proof of the derivation system with
  respect to a formal cost semantics guarantees the validity of the
  certificates.
  %
  \emph{Scalability} is attained by an efficient reduction of
  bound inference to a linear optimization problem that can be solved
  by off-the-shelf LP solvers.
  %
  The analysis framework is implemented in the publicly-available
  tool $\toolname$. An experimental evaluation demonstrates the
  advantages of the new technique with a comparison of $\toolname$ with
  existing tools on challenging micro benchmarks and the analysis of
  more than $\totalLoc$ lines of C code from the cBench benchmark suite.

@@

We have implemented the analysis system in the tool
\toolname and experimentally evaluated its effectiveness by
analyzing system code and examples from the literature.
\toolname has automatically derived global resource bounds for
more than $\totalLoc$ lines of C code from the cBench benchmark suite.
%
\iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare $\toolname$
with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.

@@

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that can
  become available during execution.
\itemskipIn
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\itemskipIn
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\itemskipIn
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\itemskipIn
\item We implemented our resource bound analysis in the publicly-available tool $\toolname$.
\itemskipIn
\item We present experiments with $\toolname$ on more
  than $\totalLoc$ lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip
%
\iffull{}
{\footnotesize \sl
[To the reviewers: The auxiliary material contains an anonymized
version of the extended article, and the source code of $\toolname$
and all examples.] }

@@

\pref{tab:cbench} contains a compilation of the results of our
experiments with the cBench benchmark suite. It shows a representative
list of automatically derived function bounds. In total we analyzed
more than $\totalLoc$ lines of code.  In the LoC column
we not only count the lines of the analyzed function but
also the ones of all the function it calls. \iffull{The line count is
computed on pre-processed C with all unnecessary declarations removed.}{}
We analyzed the functions using a metric that assigns a cost 1 to all
the back-edges in the control flow (loops, and function calls).
The bounds for the functions \code{ycc\_rgb\_conv} and \code{uv\_decode}
have been inferred with user interaction as described in \pref{sec:anno}.
The most challenging functions for $\toolname$ have unrolled loops
where many variables are assigned.  This stresses our analysis because
the number of LP variables has a quadratic growth in program variables.
Even on these stressful examples, the analysis could finish in less
than 2 seconds.  \iffull{These times can be reduced by adding
manual annotations to indicate $\toolname$ that certain variables do
not have to be tracked.  We could also envision to implement heuristics.}{}
%
For example, the \code{sha\_update} function
is composed of one loop calling two helper functions that in turn
have 6 and 1 inner loops.  In the analysis of the SHA algorithm,
the compositionality of our analysis is essential to get a tight bound
since loops on the same index are sequenced 4 and 2 times without
resetting it.  All other tools derive much larger constant factors.

@@

\begin{table}[t]
\centering
\small
\begin{tabular}{r|c|c|c}
Function & LoC & Bound & Time (s) \\
\hline
\code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
\code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
\code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
\code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
\code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
\code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
\code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
\code{MD5Final} & 195 & 141 & 0.22 \\
\code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
\code{build\_ycc\_rgb} & 68 & 257 & 0.05 \\
\code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
\code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
\hline
\code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
\code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
\end{tabular}
\vspace{.1cm}
\caption{Derived bounds for functions from cBench.}
\label{tab:cbench}
\end{table}

% \begin{table}[t]
% \small
% \centering
% \begin{tabular}{r|c|c|c}
% Function & LoC & Bound & Time (s) \\
% \hline
% \code{adpcm\_coder} & 145 & $1+|[0,\code{N}]|$ & 0.6 \\
% \code{adpcm\_decod} & 130 & $1+|[0,\code{N}]|$ & 0.2 \\
% \code{BF\_cfb64\_enc} & 151 & $1+2|[-1,\code{N}]|$ & 0.7 \\
% \code{BF\_cbc\_enc} & 180 & $2+0.25|[-8,\code{N}]|$ & 1.0 \\
% \code{mad\_bit\_crc} & 145 & $61.19{+}0.19|[-1,\code{N}]|$ & 0.4 \\
% \code{mad\_bit\_read} & 65 & $1+0.12|[0,\code{N}]|$ & 0.05 \\
% \code{MD5Update} & 200 & $133.95{+}1.05|[0,\code{N}]|$ & 1.0 \\
% \code{sha\_update} & 98 & $2 + 3.55|[0,\code{N}]|$ & 1.2 \\
% \code{PackBitsDecode} & 61 & $1+65|[-129,\code{cc}]|$ & 0.6 \\
% \code{KMPSearch} & 20 & $1+2|[0,n]|$ & 0.1 \\
% \hline
% \code{ycc\_rgb\_conv} & 66 & $\code{nr}\cdot\code{nc}$ & 0.1 \\
% \code{uv\_decode} & 31 & $\log_2(\code{UV\_NVS})+1$ & 0.1 \\
% \end{tabular}
% \vspace{.1cm}
% \caption{Derived bounds for functions from cBench.}
% \label{tab:cbench}
% \end{table}
